How predictable is extinction?

### 2019-01-04 doc/methods.tex
citations to methods

A taxon's geographic range was calculated for each of the 1 My bins in which it occurred. Geographic range was calculated as the maximum great circle distance on an ellipsoid (i.e. the globe) between any two occurrences; this measure is also called a geodesic. This distance was measured in kilometers. The geodesic distance was then log-plus-one transformed and then standardized by zero-centering the data and then dividing by its standard deviation so that geographic range had mean 0 and standard deviation 1. This standardization means that a regression coefficient associated with this covariate describes the change in extinction probability per change in standard deviation of geographic range, and that coefficients associated with similarly standardized covariates will be directly comparable in magnitude.

Temperature data used as covariates in this analysis are based on Magnesium/Calcium isotope ratios sourced from Cramer et al CITATION. These elemental ratios are considered more accurate estimates of past global temperature when compared to the frequently used Oxygen isotope-based estimates; this is because Mg/Ca based estimates are not effected by ice-volume and fresh-water input (e.g. meteoric water) which can alter Oxygen isotope ratios without reflecting changes to the climate itself. This property is of particular importance for this analysis as polar ice-caps develop midway through the Cenozoic. Our data source, Cramer et al., provides temperature estimates for every 0.1 My from 0 to 63 Mya. We binned these estimates into 1 My intervals as we did with the fossil occurrences. The temperature estimate for each 1 My interval was calculated as the mean of all estimates within that interval (Fig. \ref{fig:temp_curve}). Temperature was then transformed and standardized the in the same manner as geographic range (above).

@@

The hazard function (Eq. \ref{eq:hazard}) is easily reparameterized as a logistic regression by defining that \(\lambda(t | X) = h(\Theta)\) where \(h(.)\) is a logit inverse-link function and \(\Theta\) is the probability of a taxon going extinction during interval \(t\). \(h(\Theta)\) is then modeled as with any regression as it is defined for all real-values. In this case, we opted for a hierarchical/mixed-effects model with multiple non-nested varying intercepts and slopes CITATION.

@@

Instead of assigning a prior distribution for each of the covariance matrices in the model, we instead decomposed the covariance matrices (e.g. \(\Sigma_{B}\)) which allows us to assign independent priors for the scale and correlation aspects of covariance. The scale parameters were assigned half-Cauchy priors as described above in the context of all other scale parameters. The correlation matrices were assigned LKJ priors each with shape parameter set to 1. This choice of shape parameter produces a uniform distribution over possible correlation matrices. These priors are also slightly more interpretable than other common prior distributions for covariance matrices such as the inverse-Wishart distribution. This approach to assigning priors to a covariance matrix is recommended by CITATION.

@@

We implemented our model using the \begin{texttt}rstanarm\end{texttt} package for the R programming language CITATION. This package provides an interface to the Stan probabilistic programming language for writing hierarchical/mixed-effects models in native R. Posterior estimates were obtained through Hamiltonian Monte Carlo, using 2000 steps divided equally between warm-up and sampling. In order to prevent divergent transitions the adapt delta value was increased to 0.9999; all other HMC/NUTS sampling parameters were kept at the defaults for rstanarm version XX CITATION.

@@

Similarly, our full model can be implemented in \begin{texttt}brms\end{texttt} CITATION as:
\begin{verbatim}
priors <- c(set_prior('normal(0, 10)', class= 'Intercept'),
            set_prior('normal(0, 1)', class = 'b'),
            set_prior('normal(-1, 1)', class = 'b', coef = 'range'),
            set_prior('cauchy(0, 3)', class = 'sd'),
            set_prior('lkj(1)', class = 'cor'))
brmfit <- brm(formula = bf(event ~ range + range_diff + temp + temp_lag +
                           (1 + range + range_diff + temp + temp_lag | mybin/phylum) +
                           (1 | age/phylum))
              data = data, 
              family = bernoulli(), 
              prior = priors,
              chains = 4, 
              thin = 4,
              control = list(adapt_delta = 0.999999)
\end{verbatim}

Posterior convergence was determined using the general and HMC-specific diagnostic criteria: scale reduction factor (\(\hat{R}\); target \(<1.1\)), effective sample size (eff; target value eff/steps \(<0.0001\)), number of samples that saturated the maximum trajectory length for avoiding infinite loops (treedepth; target value 0), sample divergence, and the energy Bayesian Fraction of Mission Information (E-BFMI; target value \(>0.2\)). For further explanation of these diagnostic criteria, see the Stan Manual CITATION.

@@

See our code repository LINK for full code details. Our code uses ``tidyverse'' tools such as dplyr CITATION, purr CITATION, and tidybayes CITATION, thus some familiarity with that package ecosystem is necessary to fully comprehend how we've processed our data and results.

>>>>

A taxon's geographic range was calculated for each of the 1 My bins in which it occurred. Geographic range was calculated as the maximum great circle distance on an ellipsoid (i.e. the globe) between any two occurrences; this measure is also called a geodesic. This distance was measured in kilometers. The geodesic distance was then log-plus-one transformed and then standardized by zero-centering the data and then dividing by its standard deviation so that geographic range had mean 0 and standard deviation 1. This standardization means that a regression coefficient associated with this covariate describes the change in extinction probability per change in standard deviation of geographic range, and that coefficients associated with similarly standardized covariates will be directly comparable in magnitude \citep{ARM}.

Temperature data used as covariates in this analysis are based on Magnesium/Calcium isotope ratios sourced from \citet{Cramer2011}. These elemental ratios are considered more accurate estimates of past global temperature when compared to the frequently used Oxygen isotope-based estimates; this is because Mg/Ca based estimates are not effected by ice-volume and fresh-water input (e.g. meteoric water) which can alter Oxygen isotope ratios without reflecting changes to the climate itself. This property is of particular importance for this analysis as polar ice-caps develop midway through the Cenozoic. Our data source, \citet{Cramer2011}, provides temperature estimates for every 0.1 My from 0 to 63 Mya. We binned these estimates into 1 My intervals as we did with the fossil occurrences. The temperature estimate for each 1 My interval was calculated as the mean of all estimates within that interval (Fig. \ref{fig:temp_curve}). Temperature was then transformed and standardized the in the same manner as geographic range (above).

@@

The hazard function (Eq. \ref{eq:hazard}) is easily reparameterized as a logistic regression by defining that \(\lambda(t | X) = h(\Theta)\) where \(h(.)\) is a logit inverse-link function and \(\Theta\) is the probability of a taxon going extinction during interval \(t\). \(h(\Theta)\) is then modeled as with any regression as it is defined for all real-values. In this case, we opted for a hierarchical/mixed-effects model with multiple non-nested varying intercepts and slopes \citep{ARM}.

@@

Instead of assigning a prior distribution for each of the covariance matrices in the model, we instead decomposed the covariance matrices (e.g. \(\Sigma_{B}\)) which allows us to assign independent priors for the scale and correlation aspects of covariance. The scale parameters were assigned half-Cauchy priors as described above in the context of all other scale parameters. The correlation matrices were assigned LKJ priors each with shape parameter set to 1. This choice of shape parameter produces a uniform distribution over possible correlation matrices. These priors are also slightly more interpretable than other common prior distributions for covariance matrices such as the inverse-Wishart distribution. This approach to assigning priors to a covariance matrix is recommended by the Stan Manual \citep{StanManual}.

@@

We implemented our model using the \begin{texttt}rstanarm\end{texttt} package for the R programming language \citep{StanManual}. This package provides an interface to the Stan probabilistic programming language for writing hierarchical/mixed-effects models in native R. Posterior estimates were obtained through Hamiltonian Monte Carlo, using 2000 steps divided equally between warm-up and sampling. In order to prevent divergent transitions the adapt delta value was increased to 0.9999; all other HMC/NUTS sampling parameters were kept at the defaults for rstanarm 2.18.2 \citep{rstanarm}.

@@

Similarly, our full model can be implemented in \begin{texttt}brms\end{texttt} \citep{brms2017,brms2018} as:
\begin{verbatim}
priors <- c(set_prior('normal(0, 10)', class= 'Intercept'),
            set_prior('normal(0, 1)', class = 'b'),
            set_prior('normal(-1, 1)', class = 'b', coef = 'range'),
            set_prior('cauchy(0, 3)', class = 'sd'),
            set_prior('lkj(1)', class = 'cor'))
brmfit <- brm(formula = bf(event ~ range + range_diff + temp + temp_lag +
                           (1 + range + range_diff + temp + temp_lag | mybin/phylum) +
                           (1 | age/phylum))
              data = data, 
              family = bernoulli(), 
              prior = priors,
              chains = 4, 
              thin = 4,
              control = list(adapt_delta = 0.999999)
\end{verbatim}

Posterior convergence was determined using the general and HMC-specific diagnostic criteria: scale reduction factor (\(\hat{R}\); target \(<1.1\)), effective sample size (eff; target value eff/steps \(<0.0001\)), number of samples that saturated the maximum trajectory length for avoiding infinite loops (treedepth; target value 0), sample divergence, and the energy Bayesian Fraction of Mission Information (E-BFMI; target value \(>0.2\)). For further explanation of these diagnostic criteria, see the Stan Manual \citep{StanManual}.

@@

See our code repository LINK for full code details. Our code uses ``tidyverse'' tools such as dplyr \citep{dplyr}, purr \citep{purrr}, and tidybayes \citep{tidybayes}, thus some familiarity with that package ecosystem is necessary to fully comprehend how we've processed our data and results.
